---
title: "IARR presentation supplement"
format: revealjs
editor: visual
---

```{r warning = FALSE, error = FALSE, message = FALSE, echo=FALSE, results='hide'}
# Loading libraries
library(here)
library(tidyverse)
library(gtsummary)
library(gghighlight)
library(scales)
library(jsonlite)
library(ggrepel) 
library(patchwork)
library(MASS)
library(srvyr)
library(modelsummary)
library(ggeffects)
library(flextable)
library(ftExtra)

# Define color palette
c_palette <- c("#18BC9C",
               "#3498DB",
               "#F39C12",
               "#9966FF",
               "#E74C3C")

# Open processed data
load(file.path("data/df.Rda"))

## Create survey data 
mtf_svy <- data |>
  select(c(gdsp, gdpa, gdwk, gdsp_v, gdpa_v, gdwk_v, gdsp_num, gdpa_num, gdwk_v, 
           svyweight, year, year.c, decade,
           sex, momed, race, region,
           getmar, mar3, mardum, markids, kids3, kidsdum)) |>
  # weight data
  as_survey_design(
    ids = NULL,
    weights = svyweight)

data$mar3  <- relevel(data$mar3,  ref = 1)
data$kids3 <- relevel(data$kids3, ref = 3)

```

# Robustness Check

##  {data-menu-title="Expectations Trends"}

```{r warning = FALSE, fig.width=14, fig.asp=.6}
p_mardum_sex <- mtf_svy %>%
  filter(!is.na(mardum)) %>%
  group_by(year, sex, mardum) %>%
  summarize(vals  = survey_mean(vartype = "ci"))

p_kidsdum_sex <- mtf_svy %>%
  filter(!is.na(kidsdum)) %>%
  group_by(year, sex, kidsdum) %>%
  summarize(vals  = survey_mean(vartype = "ci"))

p_mardum_sex$cat   <- "get married" 
p_kidsdum_sex$cat  <- "very likely have children \n (if married)" 

colnames(p_mardum_sex)[colnames(p_mardum_sex)=="mardum"]    <- "response"
colnames(p_kidsdum_sex)[colnames(p_kidsdum_sex)=="kidsdum"] <- "response"

df_prop_sex <- rbind(p_mardum_sex, p_kidsdum_sex)

exp_max.sex <- df_prop_sex %>%
  group_by(cat, response) %>%
  slice_max(year, n = 1) %>%
  filter(response == 1)

exp_max.sex$label <- paste(scales::percent(exp_max.sex$vals, 1), exp_max.sex$sex)

exp_min.sex <- df_prop_sex %>%
  group_by(cat, response) %>%
  slice_min(year, n = 1) %>%
  filter(response == 1)

labels_year <- c("'76", "'92", "'08", "'23")

# tidy competency trends to overlay
df_avg2 <- df_avg

df_avg2$sex <- "All"

df_avg2$response[df_avg2$cat == 'Spouse'] <- "'very good' \n spouse"
df_avg2$response[df_avg2$cat == 'Parent'] <- "'very good' \n parent"

df_avg2$cat[df_avg2$cat == 'Spouse'] <- 'get married'
df_avg2$cat[df_avg2$cat == 'Parent'] <- 'very likely have children \n (if married)'

exp_min.df_avg2 <- df_avg2 %>%
  group_by(cat) %>%
  slice_min(year, n = 1) %>%
  filter(cat != "Worker")

```

```{r warning = FALSE, fig.width=14, fig.asp=.6}

df_prop_sex %>%
  filter(response == 1) %>%
  ggplot(aes(x = year, y = vals, ymin = vals_low, ymax = vals_upp,
             colour = sex, shape = sex)) +
  geom_line(linewidth = 2, alpha = .7) +
  geom_errorbar(alpha = .2) +
  geom_point(size = 3) +
  facet_wrap(~cat) +
  geom_text_repel(aes(label = label), data = subset(exp_max.sex, sex == "Men"), 
                  size = 7, nudge_x = 5, nudge_y = .05, segment.color = 'transparent') +
  geom_text_repel(aes(label = label), data = subset(exp_max.sex, sex == "Women"), 
                  size = 7, nudge_x = 5, nudge_y = -.05, segment.color = 'transparent') +
  geom_text_repel(aes(label = scales::percent(vals, 1)), data = subset(exp_min.sex, sex == "Men"), 
                  size = 5, nudge_y = -.07, segment.color = 'transparent') +
  geom_text_repel(aes(label = scales::percent(vals, 1)), data = subset(exp_min.sex, sex == "Women"), 
                  size = 5, nudge_y = .05, segment.color = 'transparent') +
  ## overlay competency trends
  geom_line(data = subset(df_avg2, cat != "Worker"), aes(x = year, y = vals), linewidth = 3, alpha = .6) +
  geom_text_repel(aes(label = response), data = exp_min.df_avg2, 
                  size = 5, nudge_y = -.07, segment.color = 'transparent') +
  # make it pretty
  theme_minimal(30) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), 
                     limits = c(0, 1)) +
  scale_x_continuous(limits = c(1975, 2030), breaks = c(1976, 1992, 2008, 2025), labels = labels_year) +
  scale_color_manual(values = c("#3498DB", "#E74C3C", "grey80")) +
  theme(
      plot.title          = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text.x        = element_text(face = "bold"),
      legend.position     = "none",
      panel.grid.minor    = element_blank(),
      panel.spacing       = unit(1.1, "cm", data = NULL),
      plot.caption        = element_text(color = "grey80", size = 16)) +
  labs(
    title    = "% of U.S. high school seniors who think they will _________",
    x        = NULL,
    y        = NULL,
    color    = NULL,
    caption  = "Monitoring the Future 12th Grade Surveys (1976-2023)")


```

:::: {.content-visible when-profile="speaker"}
::: notes
Know what you're thinking: share of youth who expect to take on these roles likely has changed.\
Really steady proportions until the last decade.\
dramatic drops for women as spouse and parent.\
Some decline for young men.
:::
::::

##  {data-menu-title="Adjusted for expectations"}

Models adjusted for marital and parenthood expectations

```{r warning = FALSE}

## Models with expectations as controls
polr2.sp.exp  <- polr(gdsp ~ year.c * sex + I(year.c^2) * sex + mar3 +
                       momed + race + region,
                     data = data, weights = svyweight, Hess = T)

polr2.pa.exp  <- polr(gdpa ~ year.c * sex + I(year.c^2) * sex  + kids3 +
                       momed + race + region,
                     data = data, weights = svyweight, Hess = T)

## Turn into tidy dataframes

tidySP     <- broom::tidy(polr2.sp)
tidyPA     <- broom::tidy(polr2.pa)
tidySP.exp <- broom::tidy(polr2.sp.exp)
tidyPA.exp <- broom::tidy(polr2.pa.exp)

## Transform output
tidySP <- tidySP %>%
  mutate(z_scores = estimate/std.error,
         p.value  = round(2 * (1 - pnorm(abs(z_scores))), 3),
         estimate = case_when(
           coef.type == "coefficient" ~ exp(estimate),
           coef.type == "scale"       ~ estimate))

tidyPA <- tidyPA %>%
  mutate(z_scores = estimate/std.error,
         p.value  = round(2 * (1 - pnorm(abs(z_scores))), 3),
         estimate = case_when(
           coef.type == "coefficient" ~ exp(estimate),
           coef.type == "scale"       ~ estimate))

tidySP.exp <- tidySP.exp %>%
  mutate(z_scores = estimate/std.error,
         p.value  = round(2 * (1 - pnorm(abs(z_scores))), 3),
         estimate = case_when(
           coef.type == "coefficient" ~ exp(estimate),
           coef.type == "scale"       ~ estimate))

tidyPA.exp <- tidyPA.exp %>%
  mutate(z_scores = estimate/std.error,
         p.value  = round(2 * (1 - pnorm(abs(z_scores))), 3),
         estimate = case_when(
           coef.type == "coefficient" ~ exp(estimate),
           coef.type == "scale"       ~ estimate))

## Turn into modelsummary objects
modSP            <- list(tidy = tidySP)
class(modSP)     <- "modelsummary_list"

modPA            <- list(tidy = tidyPA)
class(modPA)     <- "modelsummary_list"

modSP.exp        <- list(tidy = tidySP.exp)
class(modSP.exp) <- "modelsummary_list"

modPA.exp        <- list(tidy = tidyPA.exp)
class(modPA.exp) <- "modelsummary_list"

mods <- list(
  "full model" = modSP,
  "with expectations" = modSP.exp,
  "full model" = modPA,
  "with expectations" = modPA.exp)

 cm <- c('year.c'                             = 'Year',
         'I(year.c^2)'                        = 'Year2',
         'sexWomen'                           = "Women",
         'year.c:sexWomen'                    = 'Year * Women',
         'sexWomen:I(year.c^2)'               = 'Year2 * Women',
         'mar3Not getting married'            = 'Not getting married',
         'mar3I have no idea'                 = 'I have no idea',
         'kids3Fairly likely'                 = 'Fairly likely',
         'kids3Unlikely/Uncertain'            = 'Unlikely/Uncertain')

table <- modelsummary(mods,
                      shape = term ~ model + statistic,
                      estimate = "{estimate}{stars}",
                      gof_omit = ".*",
                      coef_map = cm,
                      fmt = fmt_decimal(digits = 3, pdigits = 3),
                      output = "huxtable") %>%
  huxtable::as_flextable() %>%
  add_body_row(values = c(" ", "coef.", "S.E.","coef.", "S.E.","coef.", "S.E.","coef.", "S.E."), colwidths = c(1,1,1,1,1,1,1,1,1)) %>%
  add_body_row(values = c(" ", "full model", "with expectations","full model", "with expectations"), colwidths = c(1,2,2,2,2)) %>%
  delete_rows(i = 3, part = "body") %>%
  add_header_row(colwidths = c(1, 4, 4), values = c(" ", "Spouse", "Parents")) %>%
   theme_vanilla() %>%
    align(i = 1, part = "header", align = "center") %>%
    align(i = 1, part = "body", align = "center") %>%
  fontsize(size = 22, part = "all")

table
```

##  {#sub-id data-menu-title="Subsample by expectations"}

```{r warning = FALSE}

## Models with subsamples
polr2.sp.mar  <- polr(gdsp ~ year.c * sex + I(year.c^2) * sex + 
                        momed + race + region,
                      data=subset(data, mar3=="Getting married"), weights = svyweight, Hess = T)

polr2.pa.kids <- polr(gdpa ~ year.c * sex + I(year.c^2) * sex  +
                        momed + race + region,
                      data=subset(data, kids3=="Very likely"), weights = svyweight, Hess = T)

## Average Predictions 
pp_sp_sex_sub   <- predict_response(polr2.sp.mar, terms = c("year.c [all]", "sex"))
pp_pa_sex_sub   <- predict_response(polr2.pa.kids, terms = c("year.c [all]", "sex"))

## Combine dfs
pp_sp_sex_sub$cat    <- "Spouse" 
pp_pa_sex_sub$cat    <- "Parent" 

pp_sp_sex_sub$sample  <- "think will 'get married'" 
pp_pa_sex_sub$sample  <- "if marry, very likely have kids" 

df_pp_sex_sub <- rbind(pp_sp_sex_sub, pp_pa_sex_sub)

## Tidy variables
df_pp_sex_sub$response.level <- factor(df_pp_sex_sub$response.level, 
                                   levels=c("Very good", 
                                            "Good", 
                                            "Fairly good", 
                                            "Not so good", 
                                            "Poor"))

## Combine dfs
df_pp_sex$sample      <- "full sample" 

df_pp_p3 <- rbind(df_pp_sex, df_pp_sex_sub)


df_pp_p3$sample <- factor(df_pp_p3$sample, 
                       levels=c(
                         "full sample",
                         "think will 'get married'",
                         "if marry, very likely have kids"))

lables_year <- c("'76", "'92", "'08", "'23")

data_max.sub <- df_pp_p3 %>%
  group_by(response.level) %>%
  slice_max(x, n = 1) %>%
  filter(group == "Men")

data_min.sub <- df_pp_p3 %>%
  group_by(response.level) %>%
  slice_min(x, n = 1) %>%
  filter(response.level == "Very good")

```

:::: r-stack
```{r warning = FALSE, fig.width=14, fig.asp=.6}

df_pp_p3 %>%
    filter(cat == "Spouse") %>%
  ggplot(aes(x = x, y = predicted, color = response.level,
             ymin = conf.low, ymax = conf.high)) +
  #  geom_errorbar(width = 0.2, color="grey80") +
  geom_line(aes(linetype = group), linewidth = 2) +
  geom_text_repel(aes(label = response.level), data = data_max.sub %>% filter(cat == "Spouse"), 
                  size = 8, nudge_x = 5, segment.color = 'transparent') +
  geom_text_repel(aes(label = group), data = data_min.sub %>% filter(cat == "Spouse"), 
                  size = 8, nudge_x = 3, nudge_y = .07, segment.color = 'transparent', color = "grey30") +
  facet_wrap(~sample, ncol = 4) +
  theme_minimal(30) +
  theme(
      plot.title          = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text.x        = element_text(face = "bold"),
      legend.position     = "none",
      panel.grid.minor    = element_blank(),
      panel.spacing       = unit(1.1, "cm", data = NULL),
      plot.caption        = element_text(color = "grey70", size = 16)) +
  scale_y_continuous(breaks = c(0., .25, .5, .75), limits=c(0, .75), labels = scales::percent) +
  scale_x_continuous(breaks=c(-20.55, -4.55, 11.45, 26.45), limits=c(-21, 35), labels = lables_year) +
  scale_color_manual(values = c("#18BC9C", "#3498DB", "#9966FF","#F39C12", "#e74c3c")) +
  labs( 
    title    = "% who think they will be ' __________' as a spouse",
#    subtitle = "for the full sample and only youth expecting to get married",
    x        = " ", 
    y        = " ",
    color    = "Response level",
    linetype = " ",
    caption  = "Monitoring the Future 12th Grade Surveys (1976-2023)") 

```

::: {.fragment fragment-index="1"}
```{r warning = FALSE, fig.width=14, fig.asp=.6}

df_pp_p3 %>%
    filter(cat == "Parent") %>%
  ggplot(aes(x = x, y = predicted, color = response.level,
             ymin = conf.low, ymax = conf.high)) +
  #  geom_errorbar(width = 0.2, color="grey80") +
  geom_line(aes(linetype = group), linewidth = 2) +
  geom_text_repel(aes(label = response.level), data = data_max.sub %>% filter(cat == "Parent"), 
                  size = 8, nudge_x = 5, segment.color = 'transparent') +
  geom_text_repel(aes(label = group), data = data_min.sub %>% filter(cat == "Parent"), 
                  size = 8, nudge_x = 0, nudge_y = .07, segment.color = 'transparent', color = "grey30") +
  facet_wrap(~sample, ncol = 4) +
  theme_minimal(30) +
  theme(
      plot.title          = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text.x        = element_text(face = "bold"),
      legend.position     = "none",
      panel.grid.minor    = element_blank(),
      panel.spacing       = unit(1.1, "cm", data = NULL),
      plot.caption        = element_text(color = "grey70", size = 16)) +
  scale_y_continuous(breaks = c(0., .25, .5, .75), limits=c(0, .75), labels = scales::percent) +
  scale_x_continuous(breaks=c(-20.55, -4.55, 11.45, 26.45), limits=c(-21, 35), labels = lables_year) +
  scale_color_manual(values = c("#18BC9C", "#3498DB", "#9966FF","#F39C12", "#e74c3c")) +
  labs( 
    title    = "% who think they will be ' __________' as a parent",
#    subtitle = "for the full sample and only youth very likely to have children",
    x        = " ", 
    y        = " ",
    color    = "Response level",
    linetype = " ",
    caption  = "Monitoring the Future 12th Grade Surveys (1976-2023)") 

```
:::
::::

# Appendix

##  {#MomEd-id data-menu-title="MomEd * Year Model"}

```{r warning = FALSE}

polr2.sp.ses  <- polr(gdsp ~ year.c * momed + I(year.c^2) * momed +
                       sex + race + region,
                     data = data, weights = svyweight, Hess = T)

polr2.pa.ses  <- polr(gdpa ~ year.c * momed + I(year.c^2) * momed  + 
                       sex + race + region,
                     data = data, weights = svyweight, Hess = T)

polr2.wk.ses  <- polr(gdwk ~ year.c * momed + I(year.c^2) * momed  + 
                       sex + race + region,
                     data = data, weights = svyweight, Hess = T)

## Average Predictions 
pp_sp_ses   <- predict_response(polr2.sp.ses, terms = c("year.c [all]", "momed"))
pp_pa_ses   <- predict_response(polr2.pa.ses, terms = c("year.c [all]", "momed"))
pp_wk_ses   <- predict_response(polr2.wk.ses, terms = c("year.c [all]", "momed"))

## Combine dfs
pp_sp_ses$cat    <- "Spouse" 
pp_pa_ses$cat    <- "Parent" 
pp_wk_ses$cat    <- "Worker" 

df_pp_ses <- rbind(pp_wk_ses, pp_sp_ses, pp_pa_ses)

## Tidy variables
df_pp_ses$response.level <- factor(df_pp_ses$response.level, 
                                   levels=c("Very good", 
                                            "Good", 
                                            "Fairly good", 
                                            "Not so good", 
                                            "Poor"))

df_pp_ses$cat <- factor(df_pp_ses$cat, 
                        levels=c("Spouse", 
                                 "Parent",
                                 "Worker"))

lables_year <- c("'76", "'92", "'08", "'23")

data_max.ses <- df_pp_ses %>%
  group_by(response.level) %>%
  slice_max(x, n = 1) %>%
  filter(group == "Completed college")

data_min.ses <- df_pp_ses %>%
  group_by(response.level) %>%
  slice_min(x, n = 1) %>%
  filter(response.level == "Very good")

```

::::: r-stack
```{r warning = FALSE, fig.width=14, fig.asp=.6}

## Draw figure

df_pp_ses %>%
  filter(cat == "Spouse") %>%
  ggplot(aes(x = x, y = predicted, color = response.level,
             ymin = conf.low, ymax = conf.high)) +
  #  geom_errorbar(width = 0.2, color="grey80") +
  geom_line(aes(linetype = group), linewidth = 2) +
  geom_text_repel(aes(label = response.level), data = data_max.ses %>% filter(cat == "Spouse"), 
                  size = 8, nudge_x = 5, segment.color = 'transparent') +
  geom_text_repel(aes(label = group), data = data_min.ses %>% filter(cat == "Spouse"), 
                  size = 8, nudge_x = 3, nudge_y = .05, segment.color = 'transparent', color = "grey30") +
  theme_minimal(30) +
  theme(
      plot.title          = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text.x        = element_text(face = "bold"),
      legend.position     = "none",
      panel.grid.minor    = element_blank(),
      panel.spacing       = unit(1.1, "cm", data = NULL),
      plot.caption        = element_text(color = "grey70", size = 16)) +
  scale_y_continuous(breaks = c(0., .25, .5, .75), limits=c(0, .75), labels = scales::percent) +
  scale_x_continuous(breaks=c(-20.55, -4.55, 11.45, 26.45), labels = lables_year) +
  scale_color_manual(values = c("#18BC9C", "#3498DB", "#9966FF","#F39C12", "#e74c3c")) +
  labs( x        = " ", 
        y        = " ",
        color    = "Response level",
        title    = "% who think they will be ' __________' as a spouse",
        subtitle = "by mother's education",
        linetype = " ")

```

::: {.fragment fragment-index="1"}
```{r warning = FALSE, fig.width=14, fig.asp=.6}
df_pp_ses %>%
  filter(cat == "Parent") %>%
  ggplot(aes(x = x, y = predicted, color = response.level,
             ymin = conf.low, ymax = conf.high)) +
  #  geom_errorbar(width = 0.2, color="grey80") +
  geom_line(aes(linetype = group), linewidth = 2) +
  geom_text_repel(aes(label = response.level), data = data_max.ses %>% filter(cat == "Parent"), 
                  size = 8, nudge_x = 5, segment.color = 'transparent') +
  geom_text_repel(aes(label = group), data = data_min.ses %>% filter(cat == "Parent"), 
                  size = 8, nudge_x = 3, nudge_y = .05, segment.color = 'transparent', color = "grey30") +
  theme_minimal(30) +
  theme(
      plot.title          = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text.x        = element_text(face = "bold"),
      legend.position     = "none",
      panel.grid.minor    = element_blank(),
      panel.spacing       = unit(1.1, "cm", data = NULL),
      plot.caption        = element_text(color = "grey70", size = 16)) +
  scale_y_continuous(breaks = c(0., .25, .5, .75), limits=c(0, .75), labels = scales::percent) +
  scale_x_continuous(breaks=c(-20.55, -4.55, 11.45, 26.45), labels = lables_year) +
  scale_color_manual(values = c("#18BC9C", "#3498DB", "#9966FF","#F39C12", "#e74c3c")) +
  labs( x        = " ", 
        y        = " ",
        color    = "Response level",
        title    = "% who think they will be ' __________' as a parent",
        subtitle = "by mother's education",
        linetype = " ")
```
:::

::: {.fragment fragment-index="2"}
```{r warning = FALSE, fig.width=14, fig.asp=.6}
df_pp_ses %>%
  filter(cat == "Worker") %>%
  ggplot(aes(x = x, y = predicted, color = response.level,
             ymin = conf.low, ymax = conf.high)) +
  #  geom_errorbar(width = 0.2, color="grey80") +
  geom_line(aes(linetype = group), linewidth = 2) +
  geom_text_repel(aes(label = response.level), data = data_max.ses %>% filter(cat == "Worker"), 
                  size = 8, nudge_x = 5, segment.color = 'transparent') +
  geom_text_repel(aes(label = group), data = data_min.ses %>% filter(cat == "Worker"), 
                  size = 8, nudge_x = 3, nudge_y = .03, segment.color = 'transparent', color = "grey30") +
  theme_minimal(30) +
  theme(
      plot.title          = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text.x        = element_text(face = "bold"),
      legend.position     = "none",
      panel.grid.minor    = element_blank(),
      panel.spacing       = unit(1.1, "cm", data = NULL),
      plot.caption        = element_text(color = "grey70", size = 16)) +
  scale_y_continuous(breaks = c(0., .25, .5, .75), limits=c(0, .75), labels = scales::percent) +
  scale_x_continuous(breaks=c(-20.55, -4.55, 11.45, 26.45), labels = lables_year) +
  scale_color_manual(values = c("#18BC9C", "#3498DB", "#9966FF","#F39C12", "#e74c3c")) +
  labs( x        = " ", 
        y        = " ",
        color    = "Response level",
        title    = "% who think they will be ' __________' as a worker",
        subtitle = "by mother's education",
        linetype = " ")
```
:::
:::::

::: aside
Predicted % from ordinal logistic regression adjusting for gender, race, & region.
:::

:::: {.content-visible when-profile="speaker"}
::: notes
:::
::::

##  {background-image="images/CAF_diagram.png" background-size="contain" data-menu-title="CAF Diagram"}

::: aside
[Bear 2019](https://journals.aom.org/doi/10.5465/amr.2016.0424){target="_blank"}
:::

##  {#gap-id background-image="images/doors.png" data-menu-title="Culture vs Structure"}

:::: {.content-visible when-profile="speaker"}
::: notes
**Second Demographic Transition**\
- suggests a shift in values from traditional to individualistic norms precedes these behaviors.\
- research focuses on cultural shifts\
- changes in ideals/aspirations

**Structural changes** - economic forces account for some behavior.\
- expectations for self\
:::
::::